---
title: "NACC T-Cog Neuropsychological Assessment"
author: "Ralph Møller Trane"
format: gfm
---

This R package provides functions to work with outcomes from the NACC T-Cog Neuropsychological Assessment. The main purpose is the creation of a summary table used when dementia diagnoses are decided on through consensus (see table below^[Note that Percentile numbers are rendered weird on this front page. For more accurate representation of the table, see the `main_table()` reference page]). 

```{r}
#| eval: false
library(wlsRedcapApp)

main_table(
  demo_data, 
  studyid = "326231g", 
  cur_date = "8/9/2021"
)
```

```{r}
#| echo: false
library(wlsRedcapApp)

home_page_table <- main_table(
  demo_data, 
  studyid = "326231g", 
  cur_date = "8/9/2021"
)

gt::as_raw_html(home_page_table)
```

## Get Started

Install this package from GitHub:

```{r}
#| eval: false
remotes::install_github("rmtrane/wlsRedcapApp")
```

The main Shiny application, which allows you to sift through summary tables for all participants can be launched by first loaded the package, then calling `wlsRedcapApp()`. 

```{r}
#| eval: false

library(wlsRedcapApp)

wlsRedcapApp()
```

This package includes a demo data set (`demo_data`), which allows you to explore the functionality of the package and in particular the Shiny app. Not that these data might seem completely made up... because they are. Visits might be out of order (as in, later visits have dates earlier than first visits), and outcomes might seem contradictory. These data were created by scrambling actual PHI data, and the sole purpose is testing and showcasing this R package. If you have access to the WLS REDCap project, you can download the full data from REDCap and choose to use these data in the Shiny app. Simply follow the instructions in the app. 


